Productivity estimation in waterfowl using a non-invasive method

### 2021-11-16 Code/Draft_ruddy_duck_productivity.Rmd
update

---
lang: en-EN
bibliography: ../../../../Biblio/JabRef/Library_OFB.bib
csl: ../../../../Biblio/JabRef/Styles/APA.csl
link-citations: true
title: "Productivity estimation in waterfowl using a non-invasive method"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr - contribution in data analyses & writing]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr - contribution in writing]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk - contribution in data collection]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr - contribution in data collection]
- Matthieu Guillemain^[Office Français de la Biodiversité, France -  matthieu.guillemain@ofb.gouv.fr - contribution in writing]
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: The response of a waterfowl population to a harvest pressure depends on its capacity to renew. The recruitment, i.e. the number of young adults reproducing for the first time, is a key indicator to describe the renewal of a population, and therefore an essential tool in population management. The productivity, defined inhere as the number of recruits produced per breeder, is even more informative because it is independent of the breeding population size and allows comparisons over time and between species. The proportion of young adults in a waterfowl population, which is a main step in estimating productivity, is often estimated from game-hunting samples. However, this proportion is not accessible in years without harvest, or with a low harvest rate. Moreover, the age-structure in the harvest samples, commonly called hunting bag, does not necessarily reflect the underlying age-structure of the population. It is often skewed towards juveniles and can lead to an overestimation of the productivity. In waterfowl, adult males usually display brighter colours than females and immatures from both sexes. This dichromatism can be characterized and monitored from count surveys, a non-invasive method. Such information can be used to infer the proportion of young adults, and consequently the productivity. In using two populations of ruddy duck, *Oxyura oxyura*, this study develops a Bayesian method to estimate the productivity from count data. The results are compared to productivity estimates from samples and are discussed by testing their consistency with the population growth rates. The results suggest that the counting method is a versatile tool to estimate the productivity if the species is monitored during the appropriate time window. 
output:
  bookdown::pdf_document2:
    latex_engine: pdflatex
    toc: false
header-includes:
  \usepackage{float}
---

@@

One defines the productivity of a breeding population as the average number of recruits produced per each breeder [@Johnson1987]. This parameter might sometimes be called "annual fertility" [@Koons2014], or "reproductive success". If one considers a species reaching its sexual maturity as early as its first year, all individuals are mature at the reproduction season. The productivity is thus the ratio of the number of young adults at the reproduction season of year $t$, i.e. the recruits, to the number of all individuals at the reproduction season of the previous year, $t - 1$.

The productivity parameter is complex because it is composed of two main sub-parameters: the reproductive success and the juvenile survival. Both sub-parameters can be split into other sub-parameters. For instance, the reproductive success is the product of the nesting success and the hatching success. The juvenile survival is the combination of the post-hatching survival and the fledging juvenile survival. According to @Blums2004, the recruitment for diving ducks is not related to the reproductive success, but depends mostly on the juvenile survival, which is driven by weather conditions. However, the results of this study does not necessarily reflect a general reality because several sources of variations of the reproductive success are controlled in the experimental plan of this study. The productivity as defined in the present study is thus a parameter integrating many sources of variability affecting all life stages from the egg to the young adult trying to breed for its first time.

@@

The management of an exploited population generally aims at preserving its conservation status, generally by targeting a population size **ref required**. To do so, a first strategy is to test iteratively management rules to reach progressively an optimum. Such approach can be time-consuming given the intrinsic variability of the population response to its changing environment. It might also suffer from a rejection of the stakeholders because of the lack of transparency of its arbitrary process **ref required** [@Nichols2000]?. Another approach is to try to predict the growth rate of the population as a function of the harvest rate and other environment inputs. Such modelling approach can provide clues on the recovering time of a population upon a variety of managment rules **ref required**. These decision-support tools are then key inputs to status on the best management rules to implement. The growth rate of a population can be simply defined as the sum of the adult survival and the productivity, see Equation \@ref(eq:GR) in Section \@ref(MM). The adult survival is generally estimated by capture mark recapture methods **ref required**. If the harvest pressure behaves as an additive mortality to the natural mortality, it is quite straightforward to predict its impact of the adult survival. The key to predict the growth rate of a population is then to estimate the average value and the variability of the productivity and to understand how this parameter is impacted by a harvest pressure.

@@

A waterfowl species that is released in a favourable habitat will naturally expand  [@Malthus1872]. The growth of such a population reflects that the productivity parameter is higher than the adult mortality rate. This expansion is limited by the carrying capacity of the habitat [@Sayre2008]. When a waterfowl species is endemic of an ecosystem and evolves in stable environmental conditions, the population size varies around the carrying capacity because the individuals compete for space or/and food [@Nummi2015], which prevents the population from expanding. This competition commonly induces a lower reproductive success or/and a higher juvenile mortality. The population is then less productive because the productivity parameter is affected by density-dependent effects [@Gunnarsson2013]. For an endemic species evolving in stable environmental conditions, a balance between the productivity and the adult mortality rate explains the stabilization of the number of individuals **ref required**.

Hunting waterfowl induces adult and juvenile mortality [@Bellrose1980]. Intuitively, one expects that such a hunted population would observe lower adult survival and productivity and then would rapidly deplete to its extinction. However, hunting waterfowl has a long history, and most of these species have persisted over time **ref required**. So paradoxically, hunting immature individuals does not always induce a decrease of productivity. A reason could be that the harvest pressure reduces the competition for space and food **ref required**. The density-dependent effects affecting the productivity in case of high population density are then reduced by the harvest mortality. In conclusion, a limited exploitation of a population is expected to increase both adult mortality and productivity. A balance between these two parameters can be theoritically reached if the population is exploited to a constant and moderate harvest rate **ref required**. The size of a newly exploited population in stable environmental conditions should thus reach a different equilibrium, which is expected to be lower than the size without exploitation **ref required**.

@@

In population mangement, a common question for a threaten or an invasive species is to determine the measure that will allow to reach a targetted population size/density in a limited time period **ref required**. For such question, the carrying capacity of the habitat is theoritically not reached. So a manager should mostly focus on collecting knowledge on the maximum productivity and its variability by removing the potential density-dependent effects occurring only for the highest theoritical densities.

@@

In waterfowl, it is generally impossible to differenciate immatures from adults without examining birds in hand. Aging individuals from samples coming from hunting bags is thus another possible approach. Such approach suffers from two caveats. First, it is impossible to explore the productivity when there is no harvest whereas it is necessary to predict what would happen if a moratorium seems required to let a population recover, or to calibrate a priori the harvest effort necessary to control an invasive species [@Smith2005]. Secondly, like any predation action, hunting is selective and provides a biased picture of the population age structure in favor of the immatures [@Bellrose1980]. The hunting bag analysis might then lead to overestimate the productivity and consequently the maximum sustainable harvest rate, with adverse consequences on population management **ref required**.

>>>>

---
lang: en-EN
bibliography: ../../../../Biblio/JabRef/Library_OFB.bib
csl: ../../../../Biblio/JabRef/Styles/APA.csl
link-citations: true
title: "Productivity estimation in waterfowl using a non-invasive method"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr - contribution in data analyses & writing]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr - contribution in writing]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk - contribution in data collection]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr - contribution in data collection]
- Matthieu Guillemain^[Office Français de la Biodiversité, France -  matthieu.guillemain@ofb.gouv.fr - contribution in writing]
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: The response of a waterfowl population to a harvest pressure depends on its capacity to renew. The recruitment, i.e. the number of young adults reproducing for the first time, is a key indicator to describe the renewal of a population, and therefore an essential tool in population management. The productivity, defined inhere as the number of recruits produced per breeder, is even more informative because it is independent of the breeding population size and allows comparisons over time and between species. The proportion of young adults in a waterfowl population, which is a main step in estimating productivity, is often estimated from game-hunting samples. However, this proportion is not accessible in years without harvest, or with a low harvest rate. Moreover, the age-structure in the harvest samples, commonly called hunting bag, does not necessarily reflect the underlying age-structure of the population. It is often skewed towards juveniles and can lead to an overestimation of the productivity. In waterfowl, adult males usually display brighter colours than females and immatures from both sexes. This dichromatism can be characterized and monitored from count surveys, a non-invasive method. Such information can be used to infer the proportion of young adults, and consequently the productivity. In using two populations of ruddy duck, *Oxyura oxyura*, this study develops a Bayesian method to estimate the productivity from count data. The results are compared to productivity estimates from samples and are discussed by testing their consistency to support the observed population growth rates. The results suggest that the counting method is a versatile tool to estimate the productivity if the species is monitored during the appropriate time window. For the relevant waterfowl species, these results argue for considering a two-category protocol in winter count surveys. 
output:
  bookdown::pdf_document2:
    latex_engine: pdflatex
    toc: false
header-includes:
  \usepackage{float}
---

@@

One defines the productivity of a breeding population as the average number of recruits produced per each breeder [@Johnson1987]. This parameter might sometimes be called "annual fertility" [@Koons2014], or "reproductive rate" [@Cooch2014], or "recruitment rate" [@Flint2015]. If one considers a species reaching its sexual maturity as early as its first year, all individuals are mature at the reproduction season. The productivity is thus the ratio of the number of young adults at the reproduction season of year $t$, i.e. the recruits, to the number of all individuals at the reproduction season of the previous year, $t - 1$.

The productivity parameter is complex because it is composed of two main sub-parameters: the reproductive success followed by the juvenile survival. Both sub-parameters can be split into other sub-parameters. For instance, the reproductive success is the product of the nesting success and the hatching success. The juvenile survival is the combination of the post-hatching survival and the fledging juvenile survival. According to @Blums2004, the recruitment for diving ducks is not related to the reproductive success, but depends mostly on the juvenile survival, which is driven by weather conditions. However, the results of this study does not necessarily reflect a general reality because several sources of variations of the reproductive success are controlled in the experimental plan of this study. The productivity as defined in the present study is thus a parameter integrating many sources of variability affecting all life stages from the egg to the young adult trying to breed for its first time [@Koons2017].

@@

Population conservation, harvesting, and control require tools in population management to reach targets such as a population size for example [@Shea1998]. To do so, a first "blind" strategy is to test iteratively management rules to reach progressively an optimum. Such approach can be time-consuming given the intrinsic variability of the population response to its changing environment. It might also suffer from a rejection of the stakeholders because of the lack of transparency of an arbitrary process [@Williams2013]. Another approach is to try to predict the growth rate of the population as a function of the harvest rate and other environment inputs. Such modelling approach can provide clues on the recovering time of a population upon a variety of managment rules [@Otis2006]. These decision-support tools are then key inputs to status on the best management rules to implement. The growth rate of a population depends on the adult survival and the productivity, see Equation \@ref(eq:GR) in Section \@ref(MM). The adult survival is generally estimated by mark recapture/recovery methods [@Lebreton2001]. If the harvest pressure behaves as an additive mortality to the natural mortality for adults [e.g. @Iverson2013], it is quite straightforward to predict its impact of the adult survival. The key to predict the growth rate of a population is then to estimate the average value and the variability of the productivity and to understand how this parameter is impacted by a harvest pressure [@Anders2005].

@@

A waterfowl species that is released in a favourable habitat will naturally expand  [@Malthus1872]. The growth of such a population reflects that the productivity parameter is higher than the adult mortality rate. This expansion is limited by the carrying capacity of the habitat [@Sayre2008]. When a waterfowl species is endemic of an ecosystem and evolves in stable environmental conditions, the population size varies around the carrying capacity because the individuals compete for space or/and food [@Nummi2015], which prevents the population from expanding. This competition commonly induces a lower reproductive success or/and a higher juvenile mortality. The population is then less productive because the productivity parameter is affected by density-dependent effects [@Gunnarsson2013]. For an endemic species evolving in stable environmental conditions, a balance between the productivity and the adult mortality rate explains the stabilization of the number of individuals [@Flint2015].

Hunting waterfowl induces adult and juvenile mortality [@Bellrose1980]. Intuitively, one expects that such a hunted population would observe lower adult survival and productivity and then would rapidly deplete to its extinction. However, hunting waterfowl has a long history, and most of these species have persisted over time [@Cooch2014]. So paradoxically, hunting immature individuals does not always induce a decrease of productivity. The hetereogeneity of quality among immature individuals might explain this by inducing a compensation process if the harvest selects mostly individuals with a low reproductive potential [@Lindberg2013; @Gimenez2017]. The productivity can even be enhanced because the harvest pressure reduces the densities and potentially the competition for space and food [@Nummi2015; @Peron2012]. In conclusion, a limited exploitation of a population observing heteregeneity and/or density-dependent effect might increase or at least not affect the productivity. A balance between adult mortality and productivity can be theoritically reached if the population is exploited to a constant and moderate harvest rate [@Tsikliras2018]. The size of a newly exploited population in stable environmental conditions should thus reach a different equilibrium, which is expected to be lower than the size without exploitation.

@@

In population mangement, a common question for a threaten or an invasive species is to determine the measure that will allow to reach a targetted population size/density in a limited time period **ref required**. For such question, the carrying capacity of the habitat is theoritically not reached. So a manager should mostly focus on collecting knowledge on the maximum productivity and its variability by removing the potential density-dependent effects occurring only for the highest theoritical densities [@Peron2013].

@@

In waterfowl, it is generally impossible to differenciate immatures from adults without examining birds in hand. Aging individuals from samples coming from hunting bags is thus another possible approach. Such approach suffers from two caveats. First, it is impossible to explore the productivity when there is no harvest whereas it is necessary to predict what would happen if a moratorium seems required to let a population recover, or to calibrate a priori the harvest effort necessary to control an invasive species [@Smith2005]. Secondly, like any predation action, hunting is selective and provides a biased picture of the population age structure in favor of the immatures [@Bellrose1980; @Fox2014]. The hunting bag analysis might then lead to overestimate the productivity and consequently the maximum sustainable harvest rate, with adverse consequences on population management ; [@Fox2014].
